State of the Practice of Lattice Boltzmann Method Software Development

### 2021-06-06 StateOfPractice/Peter-Notes/Report/Report.tex
3rd edits to Answers to R Questions in LBM Report

Developers were asked to comment on obstacles in their development process, obstacles encountered by user, and potential future obstacles with the software packages.

Several development process obstacles were noted. One developer noted that their small development team, common in LBM software packages, has a lack of time to implement new features.  Team members are almost always part of the same institute or already know each other from other projects. External contributions are rare despite many of the projects accepting them.

A lack of software development experience was noted by the developer of a third package where many of the team members on the project are domain experts. There is a steep learning curve to get these team members to contribute good quality source code. This has been somewhat addressed as the code is now generally written in a way that the developer thinks makes it as easy as possible for contributions.

The developer also noted that there is a lack of incentive and funding in academia for developing good scientific software that people end up using. There are no journals that publish such source code. There is no place for the development of good quality software. Scientific software is often developed by the researchers that temporarily use it in their own research and keep the development and use in-house. This software often does not meet the standards that would be required by external users. Sometimes the software is commercialized but it is not open source.

Documentation is important, but it's quality could be improved. There is often no time or funding for maintaining quality documentation for software that is rarely used externally. Furthermore, documentation generally does not teach the underlying metaphysics. It sometimes provides an introduction to these topics. Users would need to already be familiar with these topics or spend time referencing suitable textbooks. The documentation generally focuses on explaining how to use the software. It is not feasible for package documentation to address the underlying metaphysics topics in detail. Sometimes frequently asked questions on the underlying theory are answered in the documentation, or pointers to answers are noted.

Technically, setting up parallelization was also noted as a pain point for one of the developers, and the introduction of continuous integration by another. Software development knowledge could mitigate some pain points. As already noted, many of the developers are domain experts and not professional software developers. An interviewed developer with more extensive software development experience noted that eliminating equivalent statements using macros had helped improve the quality of their source code, specifically helping with extending code to run on both the CPU and GPU.

Difficulties with ensuring correctness were also noted by several developers. They indicated that tests are run on new additions,on  both the individual modules and on the entire system to verify continued system correctness. These tests compare the package output to known correct output using test cases. A developer commented that the amount of testing data that is needed for some cases is a problem, as free services do not offer capabilities to store and frequently process such large amounts of data and in-house testing solutions needed to be created.

A few obstacles related to users were found. Several developers noted that users sometimes try to combine LBM methods with cases incorrectly, and that this is not theoretically sound. Some users think that the packages will work out of the box to solve their cases, while the packages require a good understanding of computational fluid dynamics and significant effort in setting up each case. These packages are not like more commercial software packages, they are generally set up to solve specific research problems. While they are modifiable to solve similar problems, these modifications are not trivial. So far this has been addressed by updating the documentation to better inform users of the underlying LBM theory and package requirements. Similar issues with LBM parameters were noted by another developer. Updating the user interface to better explain explain theoretical principles, as well as test user input, was the implemented solution. As noted above, sometimes frequently asked questions on the underlying theory and on how to use the software are answered in the documentation.

One developer commented that parts of their source code had been refactored to Python partially to address usability. Python was perceived as a much more usable language that would be easy for future users and developers to learn and understand. According to this developer this was one of the biggest steps in terms of addressing usability.

A few potential future obstacles were noted. One developer said that their source code had been written with a specific application in mind and that there was too much coupling between components. This would have an impact on future modifiability and reusability when trying to extend the software. The effected code would need to be refactored.

As noted above, difficulties with ensuring future correctness could also arise. As new methods and functionality is added into the software, new test cases and test data will need to be developed and provided.

>>>>

Developers were asked to comment on obstacles in their development process, obstacles encountered by users, and potential future obstacles.

Several development process obstacles were identified. One developer noted that their small development team has a lack of time to implement new features. Small development teams are common for LBM software packages. Team members are almost always part of the same institute or already know each other from other projects. External contributions are rare despite many of the projects accepting them.

A lack of software development experience was noted by another developer. Many of the team members on their project are domain experts and there is a steep learning curve to get these team members to contribute good quality source code. It was further noted that this has been somewhat addressed, as the code has been re-written to best ensure ease of future contributions. The same developer also noted that there is a lack of incentive and funding in academia for developing widely used scientific software. There are no journals that publish such source code. There is no organization helping with the development of good quality software. Scientific software is often developed in-house by the very researchers that temporarily use it in their own research. This software has a defined user and purpose, and often does not meet the standards that would be required by external users. Such software is rarely commercialized, in which case it is not open-source.

Documentation of LBM software is noted as being important, but its quality could be improved. There is often no time or funding for maintaining quality documentation for software that is rarely used outside of the development team. Furthermore, the documentation generally only provides a shallow overview of the underlying computational fluid dynamics theory. Users would be well advised to already be familiar with these topics to or they should spend significant time referencing theory resources. The documentation instead generally focuses on explaining how to use the software. It is of course not feasible for package documentation to address the underlying metaphysics topics in detail, so it is advised that the package documentation links to resources that better explain the underlying theory. Sometimes, frequently asked questions about the underlying theory are answered in the documentation.

Setting up parallelization was also noted as a technical pain point by one of the developers, and the introduction of continuous integration by another. Software development knowledge could mitigate such pain points. As already noted, many of the developers are domain experts and not professional software developers. One of the interviewed developers that has relatively more extensive software development experience noted that eliminating equivalent statements using macros had helped improve the quality of their source code, specifically helping with reusing code to run on both the CPU and GPU.

Difficulties with ensuring correctness were also noted by several developers. They indicated that tests are run on all new source code additions, testing both individual modules and the system to verify correctness. These tests compare the package output to known correct output using test cases. A developer commented that the amount of testing data that is needed for some cases is a problem as free testing services do not offer capabilities to store and process such large amounts of data, and in-house testing solutions needed to be created to address this limitation.

A few obstacles related to users were found. Several developers noted that users sometimes try to use incorrect LBM method combinations to solve their problems. Furthermore, some users think that the packages will work out of the box to solve their cases, while the packages require both a good understanding of computational fluid dynamics and an understanding of the requirements for formulating problems in the individual packages, which can be a significant endeavor. These software packages are not like commercial software packages. They are generally set up to solve specific research problems, and are otfen primarily used by their developers. While they are modifiable to solve similar problems, these modifications are not trivial. So far this problem of on-boarding new users has been addressed by updating the documentation to better inform users of the underlying LBM theory and package requirements. Similar issues with LBM parameters were noted by another developer. Updating the user interface to better explain theoretical principles, as well as test user input for compatibility, was the implemented solution. As noted above, sometimes frequently asked questions on the underlying theory and on how to use the software are answered in the documentation.

One of the interviewed developers commented that parts of their package's source code had been refactored to Python in order to help address usability issues. Python was perceived as a much more usable language, and it would be easy for future users and developers to learn and understand the source code.

A few potential future obstacles were noted. One developer said that their source code had been written with a specific application in mind and that due to this there was too much coupling between components of the source code. This would have an impact on future modifiability and reusability when trying to extend the software, and the code would need to be refactored.

As noted above, difficulties with ensuring future correctness could also arise. As new methods and functionality is added into the software, new test cases and test data will need to be developed.
